---
title: "Theory Examples for Analysis of Microbiome Time Series"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Autocorrelation -- Illustrative Examples
Reminder from text about what autocorrelation is. Here we will show (1) autocorrelation within observations of one random walk; (2) the difference between two autocorrelated time series and their differences; (3) spurious correlation of random independent time series

```{r autocorrelation}
library(tidyverse)
## First step is to generate a random walk
nsteps<-1000 # Set number of steps in random walk
set.seed(75197) # Initializing a random seed so results are reproducible
random_walk<-cumsum(rnorm(nsteps)) # Moving randomly on a 1D axis at each step according to SN
inter_step_diffs<-random_walk-dplyr::lag(random_walk,1)
## Taking care of the difference between the first time point and the pre-first time point
inter_step_diffs[1]<-0
## Pulling out x_{t-1}
previous_t<-dplyr::lag(random_walk,1)
## Preparing to plot
autocorrelation_plot_frame<-data.frame(t=1:nsteps,
                                       walk=random_walk,
                                       diffs=inter_step_diffs,
                                       prev_step=previous_t)
## Graphical parameteres
autocorrelation_theme<-theme(panel.background=element_blank(),
                             text=element_text(size=18))
## Implementing a linear model on X_t to X_t-1 to show significance of linear correlation for autocorrelated random variables
linear_model<-lm(data=autocorrelation_plot_frame,
                 walk~prev_step)
summary(linear_model)
# Modeling based off of the differences between time steps
diff_model<-lm(data=autocorrelation_plot_frame,
                   diffs~t)
summary(diff_model)
## Generating visualizations of different autocorrelated and de-autocorrelated data
walkplot<-ggplot(autocorrelation_plot_frame,aes(x=t,y=walk))+
  geom_point()+
  autocorrelation_theme+
  xlab('t')+
  ylab('X(t)')+
  ggtitle('Random Walk')
walkplot
acplot<-ggplot(autocorrelation_plot_frame,aes(x=prev_step,y=walk))+
  geom_point()+
  autocorrelation_theme+
  xlab('X(t-1)')+
  ylab('X(t)')+
  ggtitle('Autocorrelation of Random Walk\n(R^2=97.6%)')
acplot
dtacplot<-ggplot(autocorrelation_plot_frame,aes(x=t,y=diffs))+
  geom_point()+
  autocorrelation_theme+
  ylab('X(t)-X(t-1)')+
  xlab('t')+
  ggtitle('Time Differences of Random Walk\n(R^2~0)')
dtacplot
bigplot<-gridExtra::gtable_cbind(ggplotGrob(acplot),ggplotGrob(dtacplot))
## For saving to directory
#ggsave('figure1.jpg',plot=bigplot,device='jpg',dpi='retina')

## Now it seems intuitive we wouldn't attempt to compare a time series with itself as a means of assessing correlation. However, this principle holds for all independent random walks, though to a smaller degree. Consider this
set.seed(89134)
random_walks<-apply(matrix(data=rnorm(8*100),ncol=8),2,cumsum)
## These are all random and independently generated walks
random_corrs<-cor(random_walks)
## We calculate their correlations and plot them against each other
pairs(random_walks,random_walks,main='Random Walks')
corrplot::corrplot(random_corrs)
## While not every single walk is super correlated, this sample of 8 already shows a handful of very strong correlations! 
## Meanwhile if we did the differences
step_diffs<-apply(random_walks,2,function(x) x-dplyr::lag(x,1))[-1,]
pairs(step_diffs,step_diffs,main='Time Differences')
dif_cors<-cor(step_diffs)
corrplot::corrplot(dif_cors)
## For saving plots to directory
#jpeg('fig1alt_rightpanel.jpg')
#par(mfrow=c(2,1))
corrplot::corrplot(random_corrs)
corrplot::corrplot(dif_cors)
#dev.off()
#jpeg('fig1alt_lefttop.jpg',height=240)
pairs(random_walks,random_walks)
#dev.off()
#jpeg('fig1alt_leftbot.jpg',height=240)
pairs(step_diffs,step_diffs)
#dev.off()
```
## Section 2.3 Microbiome Modeling Problems are Almost Always Underdetermined
(Introductory Text Description)

```{r determination}
## Fix all this mess
# Make formula for "true curve"
# Say y=0.5x^2+2x-1 sure
x_coords<-1:4
y_coords<-(x_coords^2)/2+2*(x_coords)-1
# Now say we were to try to fit a linear model to these data.
#  A linear model in this case will be defined as:
# y=mx+b
# X and Y are our data, so there are two parameters to estimate. If we have 2 points, we can estimate both of them
obs_set_x<-x_coords[1:2]
obs_set_y<-y_coords[1:2]
m_estimate<-(y_coords[2]-y_coords[1])/(x_coords[2]-x_coords[1])
b_estimate<-y_coords[1]-m_estimate*x_coords[1]
## Commented out lines are for saving to output
#jpeg('figure2a.jpg')
plot(obs_set_x,obs_set_y,
     xlab='X',ylab='Y',main='Linear Fit')+
lines(obs_set_x,obs_set_x*m_estimate+b_estimate)
#dev.off()
# We also could estimate different values of m and b, but we can quantify the error
# for each of those estimates given the observed data and this estimate will
# always have the lowest (error=0). Now, let's say we suspected our data 
# actually have a quadratic relationship, and we wanted to estimate that with only these
# two points. We could perhaps say:
# Our model has 3 parameters
# y=ax^2+bx+c
# We could fit our data using the same approach as the linear model:
# y2-y1=a(x2^2-x1^2)+b(x2-x1)
# Then we get:
lhs<-obs_set_y[2]-obs_set_y[1]
rhs_a<-obs_set_x[2]^2-obs_set_x[1]^2
rhs_b<-obs_set_x[2]-obs_set_x[1]
# Now we have
# lhs=a*rhs_a+b*rhs_b OR
# a*((rhs_a)-lhs)/rhs_b=-b
# Which means we can calculate an a for any b that will intersect both of these points
# Example
a<-seq(0.1,2,by=0.25)
b<--1*(rhs_a*a)+lhs
# And for c we just need to plug in one of the observed points
c<-obs_set_y[1]-a*obs_set_x[1]^2-b*obs_set_x[1]
line_plotting_data<-seq(obs_set_x[1],obs_set_x[2],by=0.01)
possible_quadratics<-sapply(line_plotting_data,function(x) (a*x^2)+(b*x)+c)
# Commented out lines are for saving to output
#jpeg('figure2b.jpg')
plot(obs_set_x,obs_set_y,xlab='X',ylab='Y',main='Quadratic Fit')+
for(i in 1:nrow(possible_quadratics)){
  lines(line_plotting_data,possible_quadratics[i,])
}
#dev.off()
# All of these parameter estimations have error=0, so how do we know which one is right?
# This is the problem of undetermined model fitting. 
```